Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-08 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: added cumulative dist discussion

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) \, P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is the union between the GR waveform
model parameters~$\bm{\theta}$, detailed in Sec.~\ref{sec:review_pSEOB}, and
the non-GR parameter of $\ell$, i.e.,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\end{equation}
%
which, we recall, set the characteristic lengthscale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov-Chain
Monte Carlo algorithm~\cite{Rover:2006ni,vanderSluys:2008qx,Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$.
%
An example of this is shown in Fig. XX.

@@

There is a second criteria we must also consider.
%
As already emphasized in Ref.~\cite{Maselli:2019mjd}, the {\sc ParSpec}
formalism is by construction perturbative. This means that the non-GR
deformation parameters are small, that is,
% $\gamma \, \delta \omega^{(j)} \ll 1$ and
% $\gamma \, \delta \tau^{(j)} \ll 1$
%
\begin{equation}
\gamma \, \delta \omega^{(j)} \ll 1 \,,
\quad \textrm{and} \quad
\gamma \, \delta \tau^{(j)} \ll 1 \, \quad \textrm{(ParSpec bound)},
\label{eq:parspec_bound}
\end{equation}
%
for all $j$ in the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions for these parameters and check if
most of their support is concentrated to a domain with values much smaller than
unity.

@@

If we find that our posterior obey both conditions~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound}, we combine individual posteriors on $\ell$ from
each event.

Suppose we have a set of $N$ GW detections $\{D_{i}\}$ described by model parameters $\bm{\lambda}$.
%
% When possible, we also combine the marginalized posterior distribution on
% $\ell$ for both event GW150914 and GW200129.
%
To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.

>>>>

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) \, P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is the union between the GR waveform
model parameters~$\bm{\theta}$, detailed in Sec.~\ref{sec:review_pSEOB}, and
the non-GR parameter of $\ell$, i.e.,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\label{eq:def_params}
\end{equation}
%
which, we recall, set the characteristic lengthscale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a Markov-Chain Monte Carlo algorithm as implemented in
\texttt{LALInferenceMCMC}~\cite{Rover:2006ni,vanderSluys:2008qx},
which is part of \texttt{LALInference} software suite~\cite{Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$.
%
An example of this is shown in Fig. XX.

@@

There is a second criteria we must also consider.
%
As already emphasized in Ref.~\cite{Maselli:2019mjd}, the {\sc ParSpec}
formalism is by construction perturbative. This means that the non-GR
deformation parameters are small, that is,
%
\begin{equation}
\gamma \, \delta \omega^{(j)} \ll 1 \,,
\quad \textrm{and} \quad
\gamma \, \delta \tau^{(j)} \ll 1 \, \quad \textrm{(ParSpec bound)},
\label{eq:parspec_bound}
\end{equation}
%
for all $j$ in the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions for these parameters and check if
most of their support is concentrated to a domain with values much smaller than
unity.

@@

If, for each event, we find that the constraints~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound} are satisfied, we further combine the individual
posteriors distribution on $\ell$, allowing us to place a stronger bound on this parameter.
%
Let us review how do we calculate this \emph{cumulative posterior distribution} on $\ell$.

% 1. Description of data and assumption o \ell being shared
Suppose we have a set of $N$ GW events $\{d_{i}\}$ that are described by
model parameters $\bm{\lambda}_{i}$, where $i$ labels each event.
%
We assume that the value of the coupling constant $\ell$, given a specific theory,
is common to all events.
%
% 2. Define joint posterior
%
The joint posterior for the parameters $\bm{\lambda}_{i}$, that describe the
data $\{d_{i}\}$, is
%
\begin{equation}
P(\ell, \{\btheta_i\} | \{d_{i}\})
= \frac{ P(\ell_i, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) }.
\label{eq:def_joint_posterior}
\end{equation}
%
where we separated $\ell$ from the GR waveform model parameters $\btheta$ [cf.~Eq.~\eqref{eq:def_params}].
%
% 3. Marginalisation to obtain the cumulative posterior
%
To obtain the cumulative posterior distribution on $\ell$, we marginalize
the joint posterior~\eqref{eq:def_joint_posterior} over
all parameters $\{ \btheta_i \}$, i.e.,
%
\begin{align}
P(\ell | \{d_i\}) &= \int \dd \{\btheta_i\} \, P(\ell, \{\btheta_i\} | \{d_{i}\}),
\nonumber \\
                  &= \int \dd \{\btheta_i\} \, \frac{ P(\ell_i, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) },
\nonumber \\
                  &= \frac{P(\ell_i)}{P(\{d_i\})} \int \dd \{\btheta_i\} \, P(\{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ),
\end{align}
%
where we have assumed that the prior on $\ell$, $P(\ell)$, is independent of
the other parameters $\{\btheta_i\}$.
%
This allow us to factor $P(\lambda, \{\btheta_i\} = P(\ell) P(\{\btheta\})$
and then move $P(\ell)$ out of the integral
%
We also moved out integral the evidence $P(\{d_i\})$, which is a normalization constant.
%
% 4. Factor the integrals in a product due to statistical independence
We can now factor the probabilities inside the integral, because each event is independent from one another:
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N}
\int \dd \btheta_i \, P(\btheta_i) \, P( d_i | \ell_i, \btheta_i ).
\end{equation}
%
% 5. Final formula
We can now identify the integral as the marginalized likelihood for each individual event
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N} P( d_i | \ell_i ),
\end{equation}
%
which we can rewritte as
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N} \frac{P(d_i)}{P(\ell_i)} P(\ell_i | d_i).
\label{eq:cumulative_dist_ell}
\end{equation}
%
using Bayes' theorem. This is our final result.

In our parameter estimation studies, we found that depending on which event and
on which theory we adopt for our parameter estimation, that we had to adjust
the prior on $\ell$.
%
The reasons are twofold. First, we found that depending on the source parameters $\btheta$,
certain large values of $\ell$ could result in unphysical waveforms in the frequency band we considered.
For instance, they could feature a ringdown multiple times longer than the inspiral-merger part.
%
Such waveform are perhaps not surprising since we are, by construction, adding corrections to the GR QNM
which \emph{delay} its damping.
%
%
In practice, we found these edge cases were problematic to deal with \texttt{LALInferenceMCMC}.
%
We addressed this issue, by reducing the prior range on $\ell$, which does not pose a problem to our parameter
estimation runs because these edge-case waveforms differ grossly from the true signal and thus have
negligible likelihood
%
However, we cannot reduce the prior on $\ell$ too much. If we do so,
to the individual posterior to rail against the prior and to result in GW
waveform which are physical.
%
Altogether, this means that the priors $P(\ell)$ are not shared between events and
they must be taken into account when calculation the cumulative posterior function~\eqref{eq:cumulative_dist_ell}.

To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.
